Cannibalism among the supermassive black holes

### 2018-04-25 paper.tex
Brought this version of paper up to date with version on Authorea.

\author{Charles Zivancev\altaffilmark{1}, Andreas H.W. K\"upper\altaffilmark{1,2}, Jeremiah P. Ostriker\altaffilmark{1,3}}
\altaffiltext{1}{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Princeton University, New Jersey, USA}
\email{Correspondence to: csz2104@columbia.edu}

@@

\section{Introduction}\label{sec:introduction}
A majority of all galaxies with masses $>10^8\msun$ at $z=0$ are believed to host super-massive black holes (SMBHs) in their cores. In a $\Lambda$CDM universe, galaxies at the high end of the galaxy-mass function are merger products of tens to hundreds of such massive galaxies. Even if only a small fraction of merging galaxies hosted an SMBH in their core at the time of merging, it is reasonable to assume that, during a Hubble time of continuous growth, these galaxies they will have substantial phases in which they contain two or more SMBHs at the same time.

Here we look at the merger history of four exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs diffuse into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs is most probable in high-mass galaxies with $10^{12}<M/\msun < 10^{13}$. Galaxies with lower masses  have too few mergers with SMBH hosting galaxies. Galaxies with higher masses tend to be too extended, making dynamical friction processes inefficient and hence failing to drive SMBHs into the host galaxy core.

This paper is organized as follows: in Section \ref{sec:methods} we are going to describe the cosmological simulation from which we use the merger history to setup our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our four exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs are driven into the core of their new host galaxies and how interaction with the host black hole leads to near-ejections or mergers. The final Section \ref{sec:conclusions} contains a discussion of the result and our conclusions.

\section{Methods}\label{sec:methods}
\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_1.png}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_187.png}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_217.png}\\
\caption{default}
\label{default}
\end{center}
\end{figure*}

@@

The merger tree data was in two primary files.  One file was centered around SMBHs at each redshift slice, giving information for the black hole's seed mass, accreted and seed mass, its host galaxy, the stellar mass in its host galaxy, and time after z=4 at which the BH entered the host galaxy (if it is not the central BH).  The other file was centered on galaxies at each redshift slice, prociding information on the galaxyies' stellar mass, dark matter mass, its central BH, and any orbiting black holes within the galaxy.

@@

\subsubsection{Galaxy background potential}
The simulations from which we obtained the merger tree data considered only dark matter distributions and were run using the Navarro-Frenk-White (NFW) profile described in \citet{1997ApJ...490..493N}.  Since our study is more interested SMBH mergers at the center of galaxies, dynamical friction, and hence baryonic matter, was of considerable importance.  Therefore, NFW background potential was less suitable for our application.

@@

For our simulations, the Jerry profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair, was used.  In order to fully parametrize the Jerry profile so as to use it in our simulations, we needed to match the stellar and halo mass data from the Kulier et al simulations, to the velocity dispersion at the galaxy's effective radius.  In Kulier's simulations, a galaxy's effective radius $R_{e}$ and velocity dispersion $\sigma(R_e)$ at the effective radius are:
\begin{equation} \label{re}
R_{e} = 2.5 kpc\left(\frac{M_*}{10^{11}M_{\odot}}\right)^{0.73}(1+z)^{-0.98}
\end{equation}
\begin{equation} \label{sig}
\sigma(R_{e}) = 190km/s\left(\frac{M_{*}}{10^{11}M_{\odot}}\right)^{0.2}(1+z)^{0.47}
\end{equation}

Given the galaxy's stellar mass $M_{*}$ at redshift z from Kulier's simulations, we can calculate the effective radius and velocity dispersion.  We can then equate these values to the analytic expressions for $\sigma(R_{e})$ in the Jerry profile.  Using a core radius $r_c$ = 100, we can solve for the halo radius $r_h$ using a simple recursive Newton method.  Whether $\sigma_{near}$ or $\sigma_{far}$ is used is determined by whether $R_e$ is less than or greater than $\sqrt{r_c r_h}$.

\subsubsection{Phase-space diffusion}
Weak encounters with background stars will let the SMBHs diffuse through phase space while they are orbiting within the gravitational potential of the galaxy.

\subsubsection{Gravitational wave recoils}
The code \textsc{AR-Chain} includes PN terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in \citet{2015ApJ...799..178K}, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}. To save computational time, we assume that a merger will be inevitable when the separation between two SBHs gets smaller than 10\,000 Schwarzschild radii. At the moment of the merger, we assume that the spin vectors of the two SBHs are randomly aligned. This results in kick velocities of up to several thousand km\,s$^{-1}$, with a median kick of $\approx 290$\,km\,s$^{-1}$. Since our simulations focus on NSC with relatively low escape velocities, this implies that a majority of the merging SBHs escape from the NSCs.

@@

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/radius_A.png}
\includegraphics[width=0.45\textwidth]{plots/radius_B.png}\\
\includegraphics[width=0.45\textwidth]{plots/radius_C.png}
\includegraphics[width=0.45\textwidth]{plots/radius_D.png}\\
\caption{default}
\label{default}
\end{center}
\end{figure*}

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_1.png}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_187.png}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_217.png}\\
\caption{default}
\label{default}
\end{center}
\end{figure*}

\subsection{Mergers}
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/masses_ABCD.png}
\caption{default}
\label{default}
\end{center}
\end{figure}

\noindent A: two mergers early on, and then two more at around 10-12 Gyr\\
B: 4 mergers across the final 8 Gyr\\
C: two mergers in the final Gyrs\\
D: ?\\

@@

\subsection{Stalling SMBHs in the halo}
\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_1.png}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_187.png}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_217.png}\\
\caption{default}
\label{default}
\end{center}
\end{figure*}

>>>>

\author{Charles Zivancev\altaffilmark{1}, Jeremiah P. Ostriker\altaffilmark{1,3}, Andreas H.W. K\"upper\altaffilmark{1,2}}
\altaffiltext{1}{Department of Astronomy, Columbia University, 550 West 120th Street, New York, NY 10027, USA}
\altaffiltext{2}{Hubble Fellow}
\altaffiltext{3}{Princeton University, New Jersey, USA}
\email{Correspondence to: csz2104@columbia.edu}

@@

\section{Introduction}\label{sec:introduction}
A majority of all galaxies with masses $>10^8\msun$ at $z=0$ are believed to host super-massive black holes (SMBHs) in their cores. In a $\Lambda$CDM universe, galaxies at the high end of the galaxy-mass function are merger products of tens to hundreds of such massive galaxies. Even if only a small fraction of merging galaxies hosted an SMBH in their core at the time of merging, it is reasonable to assume that, during a Hubble time of continuous growth, these galaxies will have substantial phases in which they contain two or more SMBHs at the same time.

Here we look at the merger history of four exemplary galaxies across the galaxy mass spectrum extracted from a cosmological simulation of hierarchical structure formation. We investigate how, after merging with incoming galaxies, SMBHs diffuse into the cores of the hosts and interact with the resident black hole. We show that gravitational interactions of multiple SMBHs are most probable in high-mass galaxies with $10^{12}<M/\msun < 10^{13}$. Galaxies with lower masses  have too few mergers with SMBH hosting galaxies. Galaxies with higher masses tend to be too extended, making dynamical friction processes inefficient and hence failing to drive SMBHs into the host galaxy core.

This paper is organized as follows: in Section \ref{sec:methods} we are going to describe the cosmological simulation from which we use the merger history to set up our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our four exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs are driven into the core of their new host galaxies and how interaction with the host black hole leads to near-ejections or mergers. The final Section \ref{sec:conclusions} contains a discussion of the result and our conclusions.

\section{Methods}\label{sec:methods}
\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_1.png}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_187.png}
\includegraphics[width=0.45\textwidth]{plots/Masses_plot_galaxy_217.png}\\
\caption{default}
\label{default1}
\end{center}
\end{figure*}

@@

The merger tree data was in two primary files.  One file was centered around SMBHs at each redshift slice, giving information for the black hole's seed mass, accreted and seed mass, its host galaxy, the stellar mass in its host galaxy, and time after z=4 at which the BH entered the host galaxy (if it is not the central BH).  The other file was centered on galaxies at each redshift slice, providing information on the galaxies' stellar mass, dark matter mass, central BH, and any orbiting black holes within the galaxy.

@@

\subsubsection{Galaxy background potential}
The simulations from which we obtained the merger tree data considered only dark matter distributions and were run using the Navarro-Frenk-White (NFW) profile described in \citet{1997ApJ...490..493N}.  Since our study is more interested SMBH mergers at the center of galaxies, dynamical friction, and hence baryonic matter, was of considerable importance.  Therefore, the NFW background potential was less suitable for our application.

@@

For our simulations, the Ostriker-Stone profile (\citet{2015ApJ...806L..28S}), which is a three-parameter potential-density pair, was used.  In order to fully parametrize this profile so as to use it in our simulations, we needed to match the stellar and halo mass data from the Kulier et al simulations, to the velocity dispersion at the galaxy's effective radius.  In Kulier's simulations, a galaxy's effective radius $R_{e}$ and velocity dispersion $\sigma(R_e)$ at the effective radius are:
\begin{equation} \label{re}
R_{e} = 2.5 kpc\left(\frac{M_*}{10^{11}M_{\odot}}\right)^{0.73}(1+z)^{-0.98}
\end{equation}
\begin{equation} \label{sig}
\sigma(R_{e}) = 190km/s\left(\frac{M_{*}}{10^{11}M_{\odot}}\right)^{0.2}(1+z)^{0.47}
\end{equation}

Given the galaxy's stellar mass $M_{*}$ at redshift z from Kulier's simulations, we can calculate the effective radius and velocity dispersion.  We can then equate these values to the analytic expressions for $\sigma(R_{e})$ in the Ostriker-Stone profile.  Using a core radius $r_c$ = 100, we can solve for the halo radius $r_h$ using a simple recursive Newton method.  Whether $\sigma_{near}$ or $\sigma_{far}$ is used is determined by whether $R_e$ is less than or greater than $\sqrt{r_c r_h}$.

\subsubsection{Phase-space diffusion}
Weak encounters with background stars will let the SMBHs diffuse through phase space while they are orbiting within the gravitational potential of the galaxy. The diffusion can be expressed as change in velocity of an SMBH by $\Delta \vec{v}$ per unit time. We can split this change into a component along the direction of motion of the SMBH, and one perpendicular to that. Following \citet{Binney08}, the diffusion coefficients can be expressed as 
\begin{eqnarray}
D[\Delta v_\parallel] & = & -\frac{4\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma^2}f(\chi),\label{eq:df}\\
D[(\Delta v_\parallel)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma}\frac{f(\chi)}{\chi},\\
D[(\Delta \vec{v}_\bot)^2] & = & \frac{4\sqrt{2}\pi G^2\rho(r)M_\bullet\ln\Lambda}{\sigma}\left[\frac{\mbox{erf}(\chi)-f(\chi)}{\chi}\right],
\end{eqnarray} 
where $\Delta v_\parallel \equiv \Delta \vec{v}\cdot\vec{v}/v$ is the velocity change in direction of motion, and $\Delta \vec{v}_\bot \equiv \Delta \vec{v} - \Delta v_\parallel \cdot\vec{v}/v$ is the velocity change perpendicular to the direction of motion. Here, $M_\bullet$ is the mass of the black hole, and $\chi = \frac{v}{\sqrt{2}\sigma(r)}$. The function $f(\chi)$ is given by 
\begin{equation}
f(\chi) = \frac{1}{2\chi^2}\left(\mbox{erf}(\chi)-\frac{2\chi}{\sqrt{\pi}}\exp\left(-\chi^2\right)\right).
\end{equation}
We approximate the factor $\Lambda$ in the Coulomb logarithm as
\begin{equation}
\Lambda = \left(\frac{M_{NSC}}{M_\bullet}\right)\left(\frac{r}{r_h}\right).
\end{equation}
We can identify Eq.~\ref{eq:df} as the dynamical friction term, that is, if we assumed $D[(\Delta v_\parallel)^2]  = D[(\Delta \vec{v}_\bot)^2]  = 0$, we would get Chandrasekhar's dynamical friction formula. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH gets sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to execute a random walk in phase space. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.
The velocity changes $\Delta v_\parallel$ and $\Delta\vec{v}_\bot$ per unit time $\Delta t$ can be computed with the above equations. Both changes are normal distributed, where the mean, $\mu$, and the variance, $\Sigma$, of the distributions are given by
\begin{eqnarray}
\mu_\parallel &=& D[\Delta v_\parallel]\Delta t,\\
\Sigma_\parallel &=& D[(\Delta v_\parallel)^2]\Delta t,\\
\mu_\bot &=& 0,\\
\Sigma_\bot &=& D[(\Delta \vec{v}_\bot)^2]\Delta t.
\end{eqnarray}
We compute the diffusion coefficients for each black hole at each time step, and modify its velocity on a Monte Carlo basis. For each time step we draw a random orientation before adding the perpendicular velocity change to the respective SBH. Hence, the SBH's modified velocity, $v_f$, is computed using
\begin{eqnarray}
\vec{v}_f &=& \vec{v}_0 + \Delta v_\parallel \hat{v}_\parallel + \Delta v_\bot \hat{v}_\bot,\\
\Delta v_\parallel &=& \mathcal{N}(\mu_\parallel, \Sigma_\parallel),\\
\Delta v_\bot &=& \mathcal{N}(\mu_\bot, \Sigma_\bot).
\end{eqnarray}
The change of energy, $\mbox{d}E_{BH}$, of the orbiting black hole due to phase-space diffusion is given back to the stellar background potential, with $\mbox{d}E = -\mbox{d}E_{BH}$. As a consequence of this energy transfer, inspiralling black holes will cause an expansion of the NSC. For this purpose we calculate the change in potential energy, $\mbox{d}W$, of the stellar system using
\begin{eqnarray}
E &=& T + W = \frac{1}{2}W,\\
\mbox{d}W &=& -2\,\mbox{d}E_{BH},
\end{eqnarray}
where we made use of the virial theorem $2T+W =0$. With this change in potential energy we can calculate a new radius for the stellar background potential at each integration step. For the Plummer sphere the new scale radius can be calculated as
\begin{equation}
a_{new} = a\left(1+\frac{32a\,\mbox{d}W}{3\pi GM_{NSC}^2}\right)^{-1}.
\end{equation}

\subsubsection{Gravitational wave recoils}
The code \textsc{AR-Chain} includes PN terms up to order 2.5. The SMBHs can therefore merge via gravitational wave emission. We include gravitational wave recoils following the prescription outlined in \citet{2015ApJ...799..178K}, which is based on the fitting formula by \citet{2012PhRvD..85h4015L}. To save computational time, we assume that a merger will be inevitable when the separation between two SBHs gets smaller than 1.0 Schwarzschild radii. At the moment of the merger, we assume that the spin vectors of the two SBHs are randomly aligned. This results in kick velocities of up to several thousand km\,s$^{-1}$, with a median kick of $\approx 290$\,km\,s$^{-1}$. Since our simulations focus on NSC with relatively low escape velocities, this implies that a majority of the merging SBHs escape from the NSCs.

@@

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/radius_A.png}
\includegraphics[width=0.45\textwidth]{plots/radius_B.png}\\
\includegraphics[width=0.45\textwidth]{plots/radius_C.png}
\includegraphics[width=0.45\textwidth]{plots/radius_D.png}\\
\caption{default}
\label{default2}
\end{center}
\end{figure*}

\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_1.png}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_187.png}
\includegraphics[width=0.45\textwidth]{plots/orbiting_bh_mass_histogram_gal_217.png}\\
\caption{default}
\label{default3}
\end{center}
\end{figure*}

\subsection{Mergers}
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/masses_ABCD.png}
\caption{default}
\label{default4}
\end{center}
\end{figure}

\noindent A: two mergers early on, and then two more at around 10-12 Gyr
B: 4 mergers across the final 8 Gyr
C: two mergers in the final Gyrs
D: ?

@@

\subsection{Stalling SMBHs in the halo}
\begin{figure*}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_1.png}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_65.png}\\
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_187.png}
\includegraphics[width=0.45\textwidth]{plots/t_at_center_histogram_gal_217.png}\\
\caption{default}
\label{default5}
\end{center}
\end{figure*}
